Unrestricted Termination and Non-Termination Arguments for Bit-Vector Programs\thanks{Supported by UK EPSRC EP/J012564/1 and ERC project 280053.}

### 2015-01-17 papers/termination/termination-esop/paper.tex
updated citation to the solver
small nudges

\begin{abstract}
%
Proving program termination is typically done by finding a well-founded
ranking function for the program states.  Existing termination
provers typically find ranking functions using either linear algebra or
templates.  As such they are often restricted to finding linear ranking
functions over mathematical integers.  This class of functions is
insufficient for proving termination of many terminating programs, and
furthermore a termination argument for a program operating on mathematical
integers does not always lead to a termination argument for the same program
operating on fixed-width machine integers.
We propose a termination analysis able 
to generate nonlinear, lexicographic ranking functions and
nonlinear recurrence sets that are correct for fixed-width machine arithmetic
and floating-point arithmetic
Our technique is based on a reduction from program \emph{termination} to 
second-order \emph{satisfaction}. We provide 
formulations for termination and non-termination in a fragment of second-order logic 
with restricted quantification which is decidable over finite domains \cite{kalashnikov}.
%existential second-order logic over finite domains is decidable \cite{kalashnikov}. 
The resulted technique is a sound and complete analysis for the termination 
of finite-state programs with fixed-width
integers and IEEE floating-point arithmetic.
%
\end{abstract}

@@

As the generation of such constraints is standard and covered by several other works \cite{DBLP:conf/pldi/GrebenshchikovLPR12,DBLP:conf/pldi/GulwaniSV08}, 
we will not provide the full algorithm, but rather illustrate it through the example in Figure~\ref{fig:environment-model}.
Full details of this construction appear in the extended version of this paper.
For the current example, the termination formula is given on the right side of 
Figure~\ref{fig:environment-model}:
%% Following the same design, let us assume that we have already proven the termination of loop $L_1$, and we are now trying to prove that $L_2$ and $L_3$ terminate, which we encode
%% as the conjunction of constraints on the right side of the figure:
$T_o$ is a summary of $L_1$ that over-approximates its transition relation;
$R_1$ and $R_2$ are ranking functions for $L_1$ and $L_2$, respectively.

@@

Let us assume that we have done some preprocessing of our program which has identified
loops, straight line code blocks and the control flow between these.  In particular,
the control flow analysis has determined which order these code blocks execute in,
and the nesting structure of the loops.

@@

\begin{proof}
%
Let $v_1, \ldots, v_k$ be the variables of the program $P$ under analysis,
and let each be $b$ bits wide.  Its state space $\mathcal{S}$ is then of
size $2^{bk}$.  A~ranking function $R: \mathcal{S} \to \mathcal{D}$ for $P$
exists iff $P$ terminates.  Without loss of generality,
$\mathcal{D}$ is a well-founded total order.  Since $R$ is injective, we have that $\|
\mathcal{D} \| \geq \| \mathcal{S} \|$.  If $\| \mathcal{D} \| > \|
\mathcal{S} \|$, we can construct a function $R': \mathcal{S} \to
\mathcal{D'}$ with $ \| \mathcal{D'} \| = \| \mathcal{S} \|$ by just setting
$R' = R|_\mathcal{S}$, i.e.~$R'$ is just the restriction of $R$ to
$\mathcal{S}$.  Since $\mathcal{S}$ already comes equipped with a natural
well ordering we can also construct $R'' = \iota \circ R'$ where $\iota:
\mathcal{D'} \to \mathcal{S}$ is the unique order isomorphism from
$\mathcal{D'}$ to $\mathcal{S}$.  So assuming that $P$ terminates, there is
some ranking function $R''$ that is just a permutation of $\mathcal{S}$.  If
the number of variables $k > 1$ then in general the ranking function will be
lexicographic with dimension $\leq k$ and each co-ordinate of the output
being a single $b$-bit value.

@@

\begin{proof}
 A proof of non-termination is a triple $\langle N, C, x_0 \rangle$ where
 \mbox{$N \subseteq \mathcal{S}$} is a (finite) recurrence set and
 $C : \mathcal{S} \to \mathcal{S}$ is a Skolem function choosing
 a successor for each $x \in N$.  $\mathcal{S}$ is finite, so by Lemma~\ref{lem:l-func} both
 $N$ and $C$ are computed by finite $\mathcal{L}$-programs and $x_0$ is just a ground term.
\end{proof}

>>>>

\begin{abstract}
%
Proving program termination is typically done by finding a well-founded
ranking function for the program states.  Existing termination
provers typically find ranking functions using either linear algebra or
templates.  As such they are often restricted to finding linear ranking
functions over mathematical integers.  This class of functions is
insufficient for proving termination of many terminating programs, and
furthermore a termination argument for a program operating on mathematical
integers does not always lead to a termination argument for the same program
operating on fixed-width machine integers.
We propose a termination analysis able 
to generate nonlinear, lexicographic ranking functions and
nonlinear recurrence sets that are correct for fixed-width machine arithmetic
and floating-point arithmetic.
Our technique is based on a reduction from program \emph{termination} to 
second-order \emph{satisfaction}. We provide 
formulations for termination and non-termination in a fragment of second-order logic 
with restricted quantification which is decidable over finite domains \cite{kalashnikov}.
%existential second-order logic over finite domains is decidable \cite{kalashnikov}. 
The resulting technique is a sound and complete analysis for the termination 
of finite-state programs with fixed-width
integers and IEEE floating-point arithmetic.
%
\end{abstract}

@@

As the generation of such constraints is standard and covered by several other works \cite{DBLP:conf/pldi/GrebenshchikovLPR12,DBLP:conf/pldi/GulwaniSV08}, 
we will not provide the full algorithm, but rather illustrate it through the example in Figure~\ref{fig:environment-model}.
%Full details of this construction appear in the extended version of this paper.
For the current example, the termination formula is given on the right side of 
Figure~\ref{fig:environment-model}:
%% Following the same design, let us assume that we have already proven the termination of loop $L_1$, and we are now trying to prove that $L_2$ and $L_3$ terminate, which we encode
%% as the conjunction of constraints on the right side of the figure:
$T_o$ is a summary of $L_1$ that over-approximates its transition relation;
$R_1$ and $R_2$ are ranking functions for $L_1$ and $L_2$, respectively.

@@

Let us assume that we have done some preprocessing of our program which has identified
loops, straight-line code blocks and the control flow between these.  In particular,
the control flow analysis has determined which order these code blocks execute in,
and the nesting structure of the loops.

@@

\begin{proof}
%
Let $v_1, \ldots, v_k$ be the variables of the program $P$ under analysis,
and let each be $b$ bits wide.  Its state space $\mathcal{S}$ is then of
size $2^{bk}$.  A~ranking function $R: \mathcal{S} \to \mathcal{D}$ for $P$
exists iff $P$ terminates.  Without loss of generality,
$\mathcal{D}$ is a well-founded total order.  Since $R$ is injective, we have that $\|
\mathcal{D} \| \geq \| \mathcal{S} \|$.  If $\| \mathcal{D} \| > \|
\mathcal{S} \|$, we can construct a function $R': \mathcal{S} \to
\mathcal{D'}$ with $ \| \mathcal{D'} \| = \| \mathcal{S} \|$ by just setting
$R' = R|_\mathcal{S}$, i.e.~$R'$ is just the restriction of $R$ to
$\mathcal{S}$.  Since $\mathcal{S}$ already comes equipped with a natural
well ordering we can also construct $R'' = \iota \circ R'$ where $\iota:
\mathcal{D'} \to \mathcal{S}$ is the unique order isomorphism from
$\mathcal{D'}$ to $\mathcal{S}$.  So assuming that $P$ terminates, there is
some ranking function $R''$ that is just a permutation of $\mathcal{S}$.  If
the number of variables $k > 1$, then in general the ranking function will be
lexicographic with dimension $\leq k$ and each co-ordinate of the output
being a single $b$-bit value.

@@

\begin{proof}
 A proof of non-termination is a triple $\langle N, C, x_0 \rangle$ where
 \mbox{$N \subseteq \mathcal{S}$} is a (finite) recurrence set and
 $C : \mathcal{S} \to \mathcal{S}$ is a Skolem function choosing
 a successor for each $x \in N$.  The state space $\mathcal{S}$ is finite, so by Lemma~\ref{lem:l-func} both
 $N$ and $C$ are computed by finite $\mathcal{L}$-programs and $x_0$ is just a ground term.
\end{proof}
